\documentclass[10pt,a4,]{article}
\usepackage[dvipsnames]{xcolor}
% \RequirePackage[l2tabu, orthodox]{nag}
\usepackage[a4paper,text={16.5cm,25.2cm},centering,margin=2.4cm]{geometry}
% \usepackage[left=1.0in,top=1.0in,right=1.0in,bottom=1.0in]{geometry}
\newcommand*{\authorfont}{\fontfamily{phv}\selectfont}
\usepackage{hyperref,amsmath,amssymb,bm,url,enumitem,dcolumn,upquote,framed,alltt,textgreek,xfrac,fixltx2e}
\usepackage[australian]{babel}
\usepackage[compact,small]{titlesec}
\setlength{\parskip}{1.2ex}
\setlength{\parindent}{0em}

\def\tightlist{}

\usepackage{ifxetex}
\ifxetex
 \usepackage{fontspec}
 \defaultfontfeatures{Ligatures=TeX} % To support LaTeX quoting style
 % \defaultfontfeatures{Ligatures=TeX,Numbers={OldStyle}}
 \setmainfont{Palatino}
 \setsansfont[Scale=MatchLowercase]{Avenir Next Condensed}
 \setmonofont[Scale=0.78,Color=RubineRed]{Bitstream Vera Sans Mono}
\else
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{lmodern}
  % \usepackage[full]{textcomp} % directly use the degree (and some other) symbol
\fi

% place after fonts; even better typesetting for improved readability:
\graphicspath{ {figure/} }
\usepackage{tabularx} % for 'tabularx' environment and 'X' column type
\usepackage{ragged2e}  % for '\RaggedRight' macro (allows hyphenation)
\usepackage{siunitx}
    \sisetup{%
        detect-mode,
        group-digits            = false,
        input-symbols           = ( ) [ ] - + < > *,
        table-align-text-post   = false,
        round-mode              = places,
        round-precision         = 3
        }
% \usepackage[font={small, sf}, labelfont=bf]{caption} % tweaking the captions
\usepackage[font={small}, labelfont=bf]{caption} % tweaking the captions
\usepackage[color=yellow, textsize=tiny]{todonotes}
\frenchspacing%

\usepackage{abstract}
\renewcommand{\abstractname}{} % clear the title
\renewcommand{\absnamepos}{empty} % originally center

\renewenvironment{abstract}
{{%
\setlength{\leftmargin}{0mm}
\setlength{\rightmargin}{\leftmargin}%
}%
\relax}
{\endlist}

\makeatletter
\def\@maketitle{%
\newpage
%  \null
%  \vskip 2em%
%  \begin{center}%
\let \footnote \thanks
 {\fontsize{18}{20}\selectfont\raggedright  \setlength{\parindent}{0pt} \@title \par}%
}
%\fi
\makeatother


\setcounter{secnumdepth}{0}





\usepackage{longtable,booktabs}
\setlength\heavyrulewidth{0.1em}
\setlength\lightrulewidth{0.0625em}


\title{Beta-diversity  }

\author{\Large AJ
Smit\vspace{0.05in} \newline\normalsize\emph{University of the Western
Cape}  }

\date{}

% PENALTIES
\widowpenalty=1000
\clubpenalty=1000
\doublehyphendemerits=9999 % Almost no consecutive line hyphens
\brokenpenalty=10000 % No broken words across columns/pages
\interfootnotelinepenalty=9999 % Almost never break footnotes

% SECTION, SUBSECETC.TITLES
\usepackage[compact]{titlesec}
\titleformat{\chapter}
  {\normalfont\LARGE\sffamily\bfseries}
  {\thechapter}
  {1em}
  {}
\titleformat{\section}
  {\normalfont\LARGE\sffamily\bfseries}
  {\thesection}
  {1em}
  {}
\titleformat{\subsection}
  {\normalfont\Large\sffamily\bfseries}
  {\thesubsection}
  {1em}
  {}
\titleformat{\subsubsection}
  {\normalfont\large\sffamily\bfseries\slshape}
  {\thesubsubsection}
  {1em}
  {}
% \titlespacing*{<command>}{<left>}{<before-sep>}{<after-sep>}
\titlespacing*{\chapter}
  {0pt}
  {1.2ex plus 1ex minus .2ex}
  {0.5ex plus .1ex minus .1ex}
\titlespacing*{\section}
  {0pt}
  {1.2ex plus 1ex minus .2ex}
  {0.5ex plus .1ex minus .1ex}
\titlespacing*{\subsection}
  {0pt}
  {1.2ex plus 1ex minus .2ex}
  {0.5ex plus .1ex minus .1ex}
\titlespacing*{\subsubsection}
  {0pt}
  {1.2ex plus 1ex minus .2ex}
  {0.5ex plus .1ex minus .1ex}




\newtheorem{hypothesis}{Hypothesis}
\usepackage{setspace}

\makeatletter

\@ifpackageloaded{hyperref}{}{%
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[colorlinks=true, citecolor=blue, linkcolor=cyan, pdfborder={0 0 0 }, unicode=true]{hyperref} % place after other packages, but before cleveref
\fi
}

\@ifpackageloaded{color}{
    \PassOptionsToPackage{usenames,dvipsnames}{color}
}{%
    \usepackage[usenames,dvipsnames]{color}
}

\usepackage{cleveref} % cleverly cross referencing figures and tables; last package to include
\setcounter{secnumdepth}{2}
\setcounter{tocdepth}{2}

% To use for resetting the numbering of Appendix Tables and Figures:
%\setcounter{table}{0}
%\renewcommand{\thetable}{A\arabic{table}}
%\setcounter{figure}{0}
%\renewcommand{\thefigure}{A\arabic{figure}}

\makeatother
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={AJ Smit (University of the Western Cape)},
             pdfkeywords = {},
            pdftitle={Beta-diversity},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls


\begin{document}

% \maketitle

{% \usefont{T1}{pnc}{m}{n}
\setlength{\parindent}{0pt}
\thispagestyle{plain}
{\fontsize{18}{20}\selectfont\raggedright
\maketitle  % title \par
}
{
   \vskip 13.5pt\relax \normalsize\fontsize{11}{12}
\textbf{\authorfont AJ Smit} \hskip 15pt \emph{\small University of the
Western Cape}   
}
}



\vskip 6.5pt

\noindent 

\section{Beta-diversity}

β-diversity is a concept that describes how species assemblages
(communities) measured within the ecosystem of interest vary from place
to place, e.g.~between the various transects or quadrats used to sample
the ecosystem. β-diversity results from habitat heterogeneity (along
gradients, or randomly). We have already seen two concepts of
β-diversity, viz.~true β-diversity and absolute species turnover -- both
of these rely on knowledge of species richness at local (a measure of
\textalpha-diversity) and regional (\textgamma-diversity) scales. Much
more insight into species assembly processes can be extracted, however,
when we view \textbeta-diversity as a dissmilarity index. In this view,
we will see that there are two processes by which β-diversity might be
affected (i.e.~in which the patterning of communities over landscapes
might arise):

\begin{description}
\item[Process 1] If a region is comprised of the species A, B, C, …, M (i.e. γ-diversity is 13), a subset of the regional flora as captured by one quadrat might be species \underline{\textbf{A}}, \underline{\textbf{D}}, E, whereas in another quadrat it might be species \underline{\textbf{A}}, \underline{\textbf{D}}, F. In this instance, the α-diversity is 3 in both instances, and heterogeneity (and hence β-diversity) results from the fact that the first quadrat has species E but the other has species F. In other words, here we have the same number of species in both quadrats, but only two of the species are the same. The process responsible for this form of β-diversity is species ‘turnover’ ($\beta_\text{sim}$). Turnover refers to processes that cause communities to differ due to species being lost and/or gained from section to section, i.e. the species composition changes between sections without corresponding changes in α-diversity.

\item[Process 2] Consider again species A, B, C, …, M. Now we have the first quadrat with species \underline{\textbf{A}}, \underline{\textbf{B}}, C, D, \underline{\textbf{G}}, H (α-diversity is 6) and the second quadrat has a subset of this, e.g. only species \underline{\textbf{A}}, \underline{\textbf{B}}, \underline{\textbf{G}} (α-diversity 3). Here, β-diversity comes from the fact that even if the two places share the same species, the number of species can still differ amongst the quadrats (i.e. from place to place) due to one quadrat capturing only a subset of species present in the other. This form of \textbeta-diversity is called ‘nestedness-resultant’ \textbeta-diversity ($\beta_\text{sne}$), and it refers to  processes that cause species to be gained or lost, and the community with the lowest α-diversity is a subset of the richer community.
\end{description}

The above two examples show that β-diversity is coupled not only with
the identity of the species in the quadrats, but also α-diversity --
with species richness in particular.

We express β-diversity as `nestedness-resultant' (β\textsubscript{sne})
and `turnover' (β\textsubscript{sim}) components so as to be able to
distinguish between these two processes. It allows us to make inferences
about the two possible drivers of β-diversity. Turnover refers to
processes that cause communities to differ due to species being lost
and/or gained from section to section, i.e.~the species composition
changes between sections without corresponding changes in α-diversity.
The nestedness-resultant component implies processes that cause species
to be gained or lost, and the community with the lowest α-diversity is a
subset of the richer community.

\end{document}
