---
title: 'Euclidian distance'
author:
- affiliation: University of the Western Cape
  name: AJ Smit
date: '2020/08/03 (updated: `r Sys.Date()`)'
output:
  pdf_document:
    fig_caption: yes
    fig_height: 4.5
    fig_width: 4.5
    highlight: default
    keep_tex: yes
    latex_engine: xelatex
    template: template-article_ajs.tex
  html_document:
    fig_caption: yes
    fig_height: 7
    fig_retina: 2
    fig_width: 7
    highlight: espresso
    theme: cerulean
    toc_float: yes
  md_document:
    variant: markdown_github
  word_document:
    fig_caption: yes
    fig_height: 7
    fig_width: 7
    highlight: zenburn
    toc: yes
fontsize: 10pt
language: Australian
csl: /Users/ajsmit/Dropbox/R/workshops/Basic_stats/LaTeX/frontiers.csl
papersize: A4
tables: yes
bibliography: bibliography.bib
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.path = 'compliled_figures/', include = TRUE, 
                      echo = TRUE, warning = FALSE, message = FALSE, 
                      tidy = FALSE, width.cutoff = 80, cache = FALSE, 
                      size = "small",
                      # results = "hide",
                      fig.align = "center",
                      fig.width = 4,
                      fig.height = 2)
```


```{r setup}
library(tidyverse)
```

\section{Definition}

The Euclidean distance between points \textbf{p} and \textbf{q} is the length of the line segment connecting them ($\overline{\mathbf{p}\mathbf{q}}$).

On an Euclidian plane, the position of point \textbf{p} is defined by the position $(p_{x},p_{y})$ and the position of point q is $(q_{x},q_{y})$. The shortest distance between the points is therefore given by the Pythagorean Theorem

```{r}
pts <- data.frame(x = c(1, 3),
                  y = c(2, 4))
rownames(pts) <- c("p", "q")

ggplot(data = pts, aes(x = x, y = y, label = rownames(pts))) +
  geom_point() +
  geom_text(check_overlap = TRUE, nudge_x = -0.05, nudge_y = 0.05) +
  geom_line(linetype = "dashed")
```


$$ d(\mathbf{p},\mathbf{q}) = \sqrt{(p_{x} - q_{x})^{2} + (p_{y} - q_{y})^2} $$ 

```{r}
ex.xyz <- read.csv("../data/Euclidian_dustance_demo_data_xyz.csv")
library(vegan)
ex.xyz.euc <- vegdist(ex.xyz[,2:4], method = "euclidian")
```

```{r}
ex.env <- read.csv("../data/Euclidian_dustance_demo_data_env.csv")
ex.env.euc <- vegdist(ex.env[,2:4], method = "euclidian")

```

